Variable Selection in Convex Function Estimation

### 2013-12-13 tex/scam/additive_faithful.tex
major proof revamp. three-points property

Suppose we have the additive approximation:
\begin{equation}
f_k^*, \mu^* \coloneqq \argmin_{f_1,\ldots, f_s, \mu} \Bigl\{ 
             \E ( f(X) - \sum_{k=1}^s f_k(X_k) -\mu )^2 
         \,:\, \E f_k(X_k) = 0 \Bigr\}.
\end{equation}
We say that $f$ has \textbf{additive recall} under $F$ if $f^*_k = 0 \Rightarrow \trm{$f$ does not depend on coordinate $k$}$.

We say that $f$ is \textbf{additively faithful} under $F$ in case $f^*_k = 0$ iff $f$ does not depend on coordinate $k$. 
\end{definition}
% We can define the support $\trm{supp}(f) \coloneqq \{ k \,:\,
% \trm{$k$ is relevant to $f$}\}$. Let $f^* = \sum_{k=1}^s$, then $f$
% is additively faith if $\trm{supp}(f) = \trm{supp}(f^*)$.

Additive faithfulness is an attractive property because it implies that, in the population setting, the additive approximation yields consistent variable selection. Additive recall is a weaker condition but still useful: we can be sure that the additive approximation is conservative, that it does not exclude important variables.

@@

Remarkably, under a general class of distributions which we characterize below, convex multivariate functions has additive recall and under product distributions, convex functions are additively faithful.

\begin{definition}
Given a conditional density $p(\mathbf{x}_{-j} \given x_j)$, we say that $\mathbf{X}_{-j}$ is \textbf{intermittently independent} of $X_j$ if the set of $x_j$ for which $\frac{\partial p(\mathbf{x}_{-j} \given x_j)}{\partial x_j} = 0$ for all $\mathbf{x}_{-j}$ is either the whole line $[0,1]$, or a collection of at least two disjoint closed intervals.

Let $F$ be a distribution on $[0,1]^s$. We say that $F$ is an \textbf{intermittently independent} distribution if $F$ has a non-zero density $p$ with the property that, for all $j=1,...,s$, $\mathbf{X}_{-j}$ is intermittently independent of $X_j$.
\end{definition}

\begin{example} There are many examples of distributions which satisfy intermittent independence:
\begin{packed_enum}
\item $F$ is a product distribution.
\item $F$ has a density $p$ that is piece-wise constant on a gridding of $[0,1]^s$.
\item $F$ has a density $p$ that is, for some $\frac{1}{2} > \epsilon > 0$,  arbitrary on $(\epsilon, 1-\epsilon)^s$ but constant elsewhere. 
\end{packed_enum}
\end{example}

\begin{theorem}
\label{thm:convex_recall}
Let $F$ be an \textbf{intermittently independent} distribution supported on $C=[0,1]^s$ with positive density $p$. If $f$ is convex and twice differentiable, then $f$ has the \emph{additive recall} property under $F$.
\end{theorem}

\begin{theorem}
\label{thm:convex_faithful}
Let $F$ be a product distribution supported on $C=[0,1]^s$ with positive density $p$. If $f$ is convex and twice differentiable, then $f$ is \emph{additively faithful} under $F$.
\end{theorem}

@@

We give the proof of Theorem~\ref{thm:convex_recall}, which also proves one direction of Theorem~\ref{thm:convex_faithful}. The other direction of Theorem~\ref{thm:convex_faithful} is trivial.

\begin{proof} (of Theorem~\ref{thm:convex_recall})\\
Fix $k$. Using the result of Lemma~\ref{lem:general_int_reduction}, we need only show that for all $x_k$, $ \E[ f(X) - \sum_{k'} f_{k'}(X_{k'}) \given x_k] - \E f(X) = 0 $ implies that $f$ does not depend on coordinate $k$.\\

Let us assume that for all $x_k$, $ \E[ f(X) - \sum_{k'} f_{k'}(X_{k'}) \given x_k] - \E f(X) = 0 $.\\

By the intermittent independence assumption, we know that there exists at least two disjoint closed intervals $R_0, R_1$ where $p(\mathbf{x}_{-k} \given x_k)$ is independent of $x_k$. That is, there exist density functions $\phi_0(\mathbf{x}_{-k})$ and $\phi_1(\mathbf{x}_{-k})$ such that $p(\mathbf{x}_{-k} \given x_k) = \phi_0(\mathbf{x}_{-k})$ for all $x_k \in R_0$ and $p(\mathbf{x}_{-k} \given x_k) = \phi_1(\mathbf{x}_{-k})$ for all $x_k \in R_1$.

Let $x^0_k$ be an interior point of $R_0$.

For every $\mathbf{x}_{-k}$, we define $g$ as the derivative of $f$ with respect to $x_k$ at the point $(\mathbf{x}_{-k}, x^0_k)$.
\[
g(\mathbf{x}_{-k}) \coloneqq  \lim_{x_k \rightarrow x^0_k} 
     \frac{f(x_k, \mathbf{x}_{-k}) - f(x^0_k, \mathbf{x}_{-k})}{x_k - x^0_k}
\]
$g(\mathbf{x}_{-k})$ is well-defined by the assumption that $f$ is everywhere differentiable.

Now, we use $r(\mathbf{x}_{-k})$ as a short-hand for $\sum_{k'\neq k} f_{k'}(x_{k'})$ and derive that for all $x_k \in R_0$, we have:
\begin{align*} 
& \int_{\mathbf{x}_{-k}}  \phi_0(\mathbf{x}_{-k}) 
   \Big( f(x_k, \mathbf{x}_{-k}) - f(x_k^0, \mathbf{x}_{-k}) \Big) d\mathbf{x}_{-k} \\
 &= \int_{\mathbf{x}_{-k}}  \phi_0(\mathbf{x}_{-k})
   \Big( \big(f(x_k, \mathbf{x}_{-k}) - r(\mathbf{x}_{-k}) - \mu^* \big) - 
         \big(f(x^0_k, \mathbf{x}_{-k}) - r(\mathbf{x}_{-k}) - \mu^* \big) \Big)
     d \mathbf{x}_{-k} \\
 &= 0
\end{align*}

The last equality follows because we assumed that $ \E[ f(X) - \sum_{k'} f_{k'}(X_{k'}) \given x_k] - \E f(X) = 0 $ for all $x_k$. \\

Since $x^0_k$ is in the interior of $R_0$, and since the domain of $f$ is a compact set, we use the bounded convergence theorem and conclude that $\int_{\mathbf{x}_{-k}} \phi_0( \mathbf{x}_{-k}) g(\mathbf{x}_{-k}) = 0$ as well.\\

We now note that $g(\mathbf{x}_{-k})$ is a component of the subgradient 
$\partial_{\mathbf{x}} f(x^0_k, \mathbf{x}_{-k})$.

Therefore, the first-order characterization of convex functions gives us that
\begin{align*}
f(x_k, \mathbf{x}_{-k}) \geq f(x^0_k, \mathbf{x}_{-k}) + g(\mathbf{x}_{-k}) (x_k - x^0_k)
   \quad \trm{for all } x_k, \mathbf{x}_{-k}
\end{align*}

Or, equivalently, for all $x_k, \mathbf{x}_{-k}$, 
\[
f(x_k, \mathbf{x}_{-k}) - f(x^0_k, \mathbf{x}_{-k}) - g(\mathbf{x}_{-k}) (x_k-x^0_k) \geq 0
\]

Now, the critical observation we make is that, for all $x_k \in R_0$, 
\[
\int_{\mathbf{x}_{-k}} \phi_0(\mathbf{x}_{-k}) 
     \Big(f(x_k, \mathbf{x}_{-k}) - f(x^0_k, \mathbf{x}_{-k}) 
     - g(\mathbf{x}_{-k}) (x_k - x^0_k) \Big) d\mathbf{x}_{-k} = 0
\]
Since $\phi_0$ is positive by assumption and the second term is non-negative, we conclude that the second term must be exactly zero, that is, for all $x_k \in R_0$, for all $\mathbf{x}_{-k}$, 
$$
f(x_k, \mathbf{x}_{-k}) = f(x^0_k, \mathbf{x}_{-k}) - g(\mathbf{x}_{-k}) (x_k - x^0_k)
$$

The Hessian of $f$ at $(x_k, \mathbf{x}_{-k})$ for $x_k \in R$ then has a zero at the $k$-th main diagonal entry. A positive semidefinite matrix with a zero on the $k$-th main diagonal entry must have only zeros on the $k$-th row and column \footnote{ See proposition 7.1.10 of \citet{HJ90}}, which means that $g(\mathbf{x}_{-k})$ must be a constant function. Since $g$ integrates to 0 under $\phi_0(\mathbf{x}_{-k})$, it must be 0.\\

Thus, $f(x_k, \mathbf{x}_{-k}) = f(x^0_k, \mathbf{x}_{-k})$ for all $x_k \in R_0$ and for all $\mathbf{x}_{-k}$.\\

We can go through the same reasoning for $R_1$  and get that $f(x_k, \mathbf{x}_{-k})$ is a constant function of $x_k$ for all $x_k \in R_1$, that is, $f(x_k, \mathbf{x}_{-k}) = f(x^1_k, \mathbf{x}_{-k})$ for all $x_k \in R_1$ for some interior point $x^1_k$ of $R_1$.

However, if $f(x_k, \mathbf{x}_{-k})$, as an one-dimensional convex function of $x_k$, is constant on two disjoint closed intervals, it must be constant for all $x_k \in [0,1]$. Therefore, for all $\mathbf{x}_{-k}$, $f(x_k, \mathbf{x}_{-k})$ is a constant as a function of $x_k$ and $f$ does not depend on $x_k$.

\end{proof}

@@

\begin{remark}
  We assume twice differentiability in
  Theorems~\ref{thm:convex_faithful} and~\ref{thm:convex_recall} to simplify the proof. 
  We believe
  this smoothness condition is not necessary because every non-smooth
  convex function can be approximated arbitrarily well by a smooth
  one.  
\end{remark}

\begin{remark}
With only intermittent independence assumption, it is difficult to prove the other direction of additive faithfulness, that is, if $f$ does not depend on coordinate $k$, then $f_k^*$ will be zero in the additive approximation. Consider as a conceptual example a 3D distribution over $(X_1, X_2, X_3)$; suppose $X_1, X_2$ are independent, and $f$ is only a function of $X_1, X_2$. We can then let $X_3 = f(X_1, X_2) - f^*_1(X_1) - f^*_2(X_2)$, that is, we let $X_3$ exactly capture the additive approximation error, then the best additive approximation of $f$ would have a component $f^*_3(X_3) = X_3$ even though $f$ does not depend on $X_3$. This is, of course, unlikely to occur in practice and we leave as future work analysis under reasonable assumptions that rules out such psychotic phenomenon. 
\end{remark}

>>>>

Suppose we have the additive approximation:
\begin{equation}
f_k^*, \mu^* \coloneqq \argmin_{f_1,\ldots, f_s, \mu} \Bigl\{ 
             \E ( f(X) - \sum_{k=1}^s f_k(X_k) -\mu )^2 
         \,:\, \E f_k(X_k) = 0 \Bigr\}.
\end{equation}

We say that $f$ is \textbf{additively faithful} under $F$ in case $f^*_k = 0 \Rightarrow \trm{$f$ does not depend on coordinate $k$}$.

\end{definition}
% We can define the support $\trm{supp}(f) \coloneqq \{ k \,:\,
% \trm{$k$ is relevant to $f$}\}$. Let $f^* = \sum_{k=1}^s$, then $f$
% is additively faith if $\trm{supp}(f) = \trm{supp}(f^*)$.

Additive faithfulness is an attractive property because it implies that, in the population setting, the additive approximation yields consistent variable selection.

@@

Remarkably, under a general class of distributions which we characterize below, convex multivariate functions are additively faithful.

\begin{definition}
\label{defn:three-point}
Let $p(\mathbf{x})$ be a density supported on $[0,1]^s$, $p$ satisfies the \emph{three-points property} if, for all $j$, and for all $\mathbf{x}_{-j}$:
\begin{packed_enum}
\item $\frac{\partial p(\mathbf{x}_{-j} \given x_j)}{\partial x_j}  = 0$ at $x_k = 0, x_k = 1$ and a third point $x_k^M \in (0,1)$.
\item $\frac{\partial^2 p(\mathbf{x}_{-j} \given x_j)}{\partial x_j^2} = 0$ at $x_k = 0, x_k = 1$.
\end{packed_enum}
\end{definition}

The three-points property is a weak condition. For instance, it is satisfied when the density is flat at the boundary of support and has at least one peak in the interior of the support. More precisely, when the \emph{joint density} satisfies the properties that $\frac{\partial p(x_j,\mathbf{x}_{-j})}{\partial x_j} = 0$ at points $x_j = 0, x_j= 1$ and a third point $x_j^M \in (0,1)$ and $\frac{\partial^2 p(x_j, \mathbf{x}_{-j})}{\partial x_j^2} = 0$ at points $x_j = 0, x_j=1$. The three-points property is also trivially satisfied when $p$ is the density of any product distributions.

The following theorem is the main result of this section.

\begin{theorem}
\label{thm:convex_faithful}
Let $p$ be a positive density supported on $C=[0,1]^s$ that satisfies the three-points property (definition~\ref{defn:three-point}). If $f$ is convex and twice differentiable, then $f$ is \emph{additively faithful} under $p$.
\end{theorem}

@@

\begin{proof} (of Theorem~\ref{thm:convex_faithful})\\
Fix $k$. Using the result of Lemma~\ref{lem:general_int_reduction}, we need only show that for all $x_k$, $ \E[ f(X) - \sum_{k'} f_{k'}(X_{k'}) \given x_k] - \E f(X) = 0 $ implies that $f$ does not depend on coordinate $k$.\\

Let us then use the shorthand notation that $r(\mathbf{x}_{-k}) = \sum_{k' \neq k} f_{k'}(x_{k'})$ and assume without loss of generality that $\mu = 0$. We then assume that for all $x_k$,

\[
 \E[ f(X) - r(X_{-k})  \given x_k] \equiv 
 \int_{\mathbf{x}_{-k}}  p(\mathbf{x}_{-k} \given x_k ) 
 \big(f(\mathbf{x}) - r(\mathbf{x}_{-k}) \big) = 0
\]

We let $p'(\mathbf{x}_{-k} \given x_k)$ denote 
$\frac{\partial p(\mathbf{x}_{-k} \given x_k)}{\partial x_k}$ and 
$p''(\mathbf{x}_{-k} \given x_k)$ denote 
$\frac{\partial^2 p(\mathbf{x}_{-k} \given x_k)}{\partial x_k^2}$ and likewise for $f'(x_k, \mathbf{x}_{-k})$ and $f''(x_k, \mathbf{x}_{-k})$. We then differentiate under the integral, which is valid because all functions are bounded.

\begin{align}
& \int_{\mathbf{x}_{-k}} p'(\mathbf{x}_{-k} \given x_k) 
\big( f(\mathbf{x}) - r(\mathbf{x}_{-k}) \big) + 
p(\mathbf{x}_{-k} \given x_k) f'(x_k, \mathbf{x}_{-k}) d \mathbf{x}_{-k}  = 0 
\label{eqn:integral1a} \\
& \int_{\mathbf{x}_{-k}} p''(\mathbf{x}_{-k} \given x_k) 
\big( f(\mathbf{x}) - r(\mathbf{x}_{-k}) \big)  + 
2 p'(\mathbf{x}_{-k} \given x_k) f'(x_k, \mathbf{x}_{-k}) +
p(\mathbf{x}_{-k} \given x_k) f''(x_k, \mathbf{x}_{-k}) d\mathbf{x}_{-k}  = 0 
\end{align}

By the three-points property, we have that $p''(\mathbf{x}_{-k} \given x_k)$ and $p'(\mathbf{x}_{-k} \given x_k)$ are zero at $x_k = x_k^0 \equiv 0$. The integral equations reduce to the following then:
\begin{align}
& \int_{\mathbf{x}_{-k}} p(\mathbf{x}_{-k} \given x^0_k) f'(x^0_k, \mathbf{x}_{-k}) d \mathbf{x}_{-k}= 0 \label{eqn:integral1b} \\
& \int_{\mathbf{x}_{-k}} p(\mathbf{x}_{-k} \given x^0_k) f''(x^0_k, \mathbf{x}_{-k}) d\mathbf{x}_{-k} = 0
\end{align}

Because $f$ is convex, $f(x_k, \mathbf{x}_{-k})$ must be a convex function of $x_k$ for all $\mathbf{x}_{-k}$. Therefore, for all $\mathbf{x}_{-k}$, $f''(x^0_k, \mathbf{x}_{-k}) \geq 0$. Since $p(\mathbf{x}_{-k} \given x^0_k) > 0$ by assumption that $p$ is a positive density, we have that $\forall \mathbf{x}_{-k}, f''(x^0_k, \mathbf{x}_{-k}) = 0$ necessarily.\\

The Hessian of $f$ at $(x^0_k, \mathbf{x}_{-k})$ then has a zero at the $k$-th main diagonal entry. A positive semidefinite matrix with a zero on the $k$-th main diagonal entry must have only zeros on the $k$-th row and column \footnote{ See proposition 7.1.10 of \citet{HJ90}}, which means that \emph{at all $\mathbf{x}_{-k}$, the gradient of $f'(x^0_k, \mathbf{x}_{-k})$ with respect to $\mathbf{x}_{-k}$ must be zero}.

Therefore, $f'(x_k^0, \mathbf{x}_{-k})$ must be a constant. By equation~\ref{eqn:integral1b}, we conclude then that $f'(x_k^0, \mathbf{x}_{-k}) = 0$ for all $\mathbf{x}_{-k}$. We can use the same reasoning for the case where $x_k = x_k^1$ and deduce that $f'(x^1_k, \mathbf{x}_{-k}) = 0$ for all $\mathbf{x}_{-k}$. \\

Now we apply the first-order condition of convex functions to the two points $(x^0_k, \mathbf{x}_{-k})$ and $(x^1_k, \mathbf{x}_{-k})$:

\begin{align*}
\forall \mathbf{x}_{-k}, f(x^1_k, \mathbf{x}_{-k}) & \leq f(x^0_k, \mathbf{x}_{-k}) 
  + f'(x^0_k, \mathbf{x}_{-k}) ( x^1_k - x^0_k) \\ 
f(x^1_k, \mathbf{x}_{-k}) &\leq f(x^0_k, \mathbf{x}_{-k}) \\
\forall \mathbf{x}_{-k}, f(x^0_k, \mathbf{x}_{-k}) & \leq f(x^1_k, \mathbf{x}_{-k}) 
  + f'(x^1_k, \mathbf{x}_{-k}) ( x^0_k - x^1_k) \\ 
f(x^0_k, \mathbf{x}_{-k}) &\leq f(x^1_k, \mathbf{x}_{-k})
\end{align*}

We thus have that $f(x^0_k, \mathbf{x}_{-k}) = f(x^1_k, \mathbf{x}_{-k})$ for all $\mathbf{x}_{-k}$. By convexity of $f$, we know that for all $x_k \in (0,1)$, it must be that for all $\mathbf{x}_{-k}$, $f(x_k, \mathbf{x}_{-k}) \leq f(x^0_k, \mathbf{x}_k)$.\\

We now use the middle point condition.

At $x_k = x^M_k$, equation~\ref{eqn:integral1a} becomes:
\begin{align}
\int_{\mathbf{x}_{-k}} p(\mathbf{x}_{-k} \given x^M_k) 
  f'(x^M_k, \mathbf{x}_{-k}) d \mathbf{x}_{-k} = 0 \label{eqn:integral1c}
\end{align}

Using the first order condition, we have that:
\begin{align*}
f(x^0_k, \mathbf{x}_{-k}) &\leq f(x^M_k, \mathbf{x}_{-k}) 
+ f'(x^M_k, \mathbf{x}_{-k})(x^0_k - x^M_k)  \quad \Rightarrow \\
\int_{\mathbf{x}_{-k}} p(\mathbf{x}_k \given x^M_0) 
f(x^0_k, \mathbf{x}_{-k}) d\mathbf{x}_{-k}  
&\leq \int_{\mathbf{x}_{-k}} p(\mathbf{x}_{-k} \given x^M_k)
\big( f(x^M_k, \mathbf{x}_{-k}) 
+ f'(x^M_k, \mathbf{x}_{-k})(x^0_k - x^M_k) \big) d\mathbf{x}_{-k} 
\end{align*}

By equation~\ref{eqn:integral1c} then, we have that $
\int_{\mathbf{x}_{-k}} p(\mathbf{x}_{-k} \given x_k) \big(f(x^0_k, \mathbf{x}_{-k}) -
 f(x^M_k, \mathbf{x}_{-k}) \big) d\mathbf{x}_{-k} \leq 0$

However, we have previously argued that for all $\mathbf{x}_{-k}$, $f(x_k, \mathbf{x}_{-k}) \leq f(x^0_k, \mathbf{x}_{-k})$ for all $x_k \in (0,1)$. Since $p(\mathbf{x}_{-k} \given x_k) > 0$, it must be that $f(x^0_k, \mathbf{x}_{-k}) = f(x^M_k, \mathbf{x}_{-k})$ for all $\mathbf{x}_{-k}$.\\

Thus, we have shown that for all $\mathbf{x}_{-k}$, 
\[
f(x^0_k, \mathbf{x}_{-k}) = f(x^M_k, \mathbf{x}_{-k}) = f(x^1_k, \mathbf{x}_{-k})
\]
This implies that $f(x_k, \mathbf{x}_{-k})$, as a convex function of $x_k$, must be a constant. This proves that $f$ does not depend on $x_k$. 
\end{proof}

@@

\begin{remark}
  We assume twice differentiability in
  Theorems~\ref{thm:convex_faithful} to simplify the proof. 
  We believe
  this smoothness condition is not necessary because every non-smooth
  convex function can be approximated arbitrarily well by a smooth
  one.  
\end{remark}

\begin{remark} 
It is difficult to prove the opposite direction of additive faithfulness, that is, if $f$ does not depend on coordinate $k$, then $f_k^*$ will be zero in the additive approximation. Consider as a conceptual example a 3D distribution over $(X_1, X_2, X_3)$; suppose $X_1, X_2$ are independent, and $f$ is only a function of $X_1, X_2$. We can then let $X_3 = f(X_1, X_2) - f^*_1(X_1) - f^*_2(X_2)$, that is, we let $X_3$ exactly capture the additive approximation error, then the best additive approximation of $f$ would have a component $f^*_3(X_3) = X_3$ even though $f$ does not depend on $X_3$. This is, of course, unlikely to occur in practice and we leave as future work analysis under reasonable assumptions that rules out such psychotic phenomenon. 
\end{remark}
